---
title: "Assignment 2"
author: "Hiba Bashir"
date: '`r Sys.Date()`'
output:
  html_document: default
  pdf_document: default
---

# Assignment 2

## Load Packages

```{r}
library(caret)
library(class)
```

## Read the Data

```{r}
bank <- read.csv("UniversalBank-1.csv")
```

## Remove ID and ZIP Code

```{r}
bank <- subset(bank, select = -c(ID, ZIP.Code))
```

## Convert Personal Loan to a Factor

```{r}
bank$Personal.Loan <- as.factor(bank$Personal.Loan)
```

## Create Dummy Variables

```{r}
dummy <- dummyVars(Personal.Loan ~ ., data = bank)

bank2 <- data.frame(
  predict(dummy, newdata = bank),
  Personal.Loan = bank$Personal.Loan
)
```

## Split the Data (60% Training / 40% Validation)

```{r}
set.seed(123)

trainIndex <- createDataPartition(
  bank2$Personal.Loan,
  p = 0.60,
  list = FALSE
)

train <- bank2[trainIndex, ]
valid <- bank2[-trainIndex, ]
```

## Normalize the Data

```{r}
normalize <- preProcess(
  train[, -ncol(train)],
  method = c("center", "scale")
)

train.x <- predict(normalize, train[, -ncol(train)])
valid.x <- predict(normalize, valid[, -ncol(valid)])

train.y <- train$Personal.Loan
valid.y <- valid$Personal.Loan
```
# Question 1

```{r}
new.customer <- as.data.frame(matrix(0, nrow = 1, ncol = ncol(train.x)))
colnames(new.customer) <- colnames(train.x)

new.customer[1, "Age"] <- 40
new.customer[1, "Experience"] <- 10
new.customer[1, "Income"] <- 84
new.customer[1, "Family"] <- 2
new.customer[1, "CCAvg"] <- 2
new.customer[1, "Mortgage"] <- 0
new.customer[1, "Securities.Account"] <- 0
new.customer[1, "CD.Account"] <- 0
new.customer[1, "Online"] <- 1
new.customer[1, "CreditCard"] <- 1
new.customer[1, "Education.1"] <- 0
new.customer[1, "Education.2"] <- 1
new.customer[1, "Education.3"] <- 0

new.customer.scaled <- predict(normalize, new.customer)
new.customer.scaled <- new.customer.scaled[, colnames(train.x)]

knn(train = train.x, test = new.customer.scaled, cl = train.y, k = 1)
```

**Answer:** Using k = 1, the customer is classified as **1**, meaning the customer is predicted to accept the personal loan offer.
# Question 2

## Determine the Best Value of k

```{r}
accuracy <- c()

for (k in 1:20) {

  pred <- knn(
    train = train.x,
    test = valid.x,
    cl = train.y,
    k = k
  )

  accuracy[k] <- mean(pred == valid.y)
}

accuracy

best.k <- which.max(accuracy)

best.k
```

**Answer:** The best value of **k** based on the validation data is **1**. This value provided the highest validation accuracy.
# Question 3

## Confusion Matrix for Validation Data

```{r}
pred.best <- knn(
  train = train.x,
  test = valid.x,
  cl = train.y,
  k = best.k
)

confusionMatrix(pred.best, valid.y, positive = "1")
```

**Answer:** The confusion matrix above shows how well the k-NN model classified the validation data using the best value of k. The model correctly predicts many customers who did not accept the loan, but the number of accepted-loan customers is smaller, so prediction for class 1 is more difficult.
# Question 4

## Classify the Customer Using the Best k

```{r}
knn(
  train = train.x,
  test = new.customer.scaled,
  cl = train.y,
  k = best.k
)
```

**Answer:** Using the best value of **k = 1**, the customer is classified as **1**, indicating that the customer is predicted to accept the personal loan offer.
# Question 5

## Repartition the Data (50% Training, 30% Validation, 20% Test)

```{r}
set.seed(123)

trainIndex2 <- createDataPartition(
  bank2$Personal.Loan,
  p = 0.50,
  list = FALSE
)

train2 <- bank2[trainIndex2, ]

remaining <- bank2[-trainIndex2, ]

validIndex2 <- createDataPartition(
  remaining$Personal.Loan,
  p = 0.60,
  list = FALSE
)

valid2 <- remaining[validIndex2, ]
test2 <- remaining[-validIndex2, ]

normalize2 <- preProcess(
  train2[, -ncol(train2)],
  method = c("center", "scale")
)

train2.x <- predict(normalize2, train2[, -ncol(train2)])
valid2.x <- predict(normalize2, valid2[, -ncol(valid2)])
test2.x <- predict(normalize2, test2[, -ncol(test2)])

train2.y <- train2$Personal.Loan
valid2.y <- valid2$Personal.Loan
test2.y <- test2$Personal.Loan

pred.train2 <- knn(train = train2.x, test = train2.x, cl = train2.y, k = best.k)
pred.valid2 <- knn(train = train2.x, test = valid2.x, cl = train2.y, k = best.k)
pred.test2 <- knn(train = train2.x, test = test2.x, cl = train2.y, k = best.k)

confusionMatrix(pred.train2, train2.y, positive = "1")

confusionMatrix(pred.valid2, valid2.y, positive = "1")

confusionMatrix(pred.test2, test2.y, positive = "1")
```

### Comparison

The training confusion matrix generally shows the best performance because the model is evaluated on the same data it learned from. The validation and test confusion matrices are more realistic measures of model performance because they use unseen data. Small differences in accuracy between the validation and test sets are expected due to differences in the observations included in each sample.
