---
title: "Assignment 3"
author: "Hiba Bashir"
output: html_document
---

## Assignment 3: Naive Bayes Classification

This assignment uses the Universal Bank data. The goal is to predict whether a customer accepts a personal loan offer using two predictors: Online banking use and Credit Card ownership.

```{r}
library(e1071)
```

## Load the Data

```{r}
bank <- read.csv(file.choose())

bank2 <- bank[, c("Online", "CreditCard", "Personal.Loan")]

names(bank2) <- c("Online", "CC", "Loan")

bank2$Online <- factor(bank2$Online)
bank2$CC <- factor(bank2$CC)
bank2$Loan <- factor(bank2$Loan)
```

## Partition the Data

The data was partitioned into 60% training data and 40% validation data.

```{r}
set.seed(1)

train.index <- sample(1:nrow(bank2), 0.6 * nrow(bank2))

train.df <- bank2[train.index, ]

valid.df <- bank2[-train.index, ]
```

## A. Pivot Table

```{r}
table(train.df$CC, train.df$Loan, train.df$Online)
```

The pivot table shows the count of customers based on credit card ownership, loan acceptance, and online banking use.

## B. Probability from Pivot Table

```{r}
subset.B <- train.df[train.df$CC == 1 & train.df$Online == 1, ]

prop.table(table(subset.B$Loan))
```

The probability that a customer who has a bank credit card and actively uses online banking accepts the loan offer is **0.0911**, or **9.11%**.

## C. Separate Pivot Tables

### Loan by Online

```{r}
table(train.df$Loan, train.df$Online)
```

### Loan by Credit Card

```{r}
table(train.df$Loan, train.df$CC)
```

## D. Conditional Probabilities

```{r}
p.cc1.loan1 <- prop.table(table(train.df$CC[train.df$Loan == 1]))["1"]

p.online1.loan1 <- prop.table(table(train.df$Online[train.df$Loan == 1]))["1"]

p.loan1 <- prop.table(table(train.df$Loan))["1"]

p.cc1.loan0 <- prop.table(table(train.df$CC[train.df$Loan == 0]))["1"]

p.online1.loan0 <- prop.table(table(train.df$Online[train.df$Loan == 0]))["1"]

p.loan0 <- prop.table(table(train.df$Loan))["0"]

p.cc1.loan1
p.online1.loan1
p.loan1
p.cc1.loan0
p.online1.loan0
p.loan0
```

The probabilities are:

1. P(CC = 1 | Loan = 1) = **0.2800**
2. P(Online = 1 | Loan = 1) = **0.6036**
3. P(Loan = 1) = **0.0917**
4. P(CC = 1 | Loan = 0) = **0.2939**
5. P(Online = 1 | Loan = 0) = **0.5828**
6. P(Loan = 0) = **0.9083**

## E. Naive Bayes Probability

```{r}
numerator <- p.cc1.loan1 * p.online1.loan1 * p.loan1

denominator <- numerator + 
  (p.cc1.loan0 * p.online1.loan0 * p.loan0)

nb.prob <- numerator / denominator

nb.prob
```

The Naive Bayes probability is **0.0906**, or **9.06%**.

## F. Comparison

The probability from the pivot table was **0.0911**, while the Naive Bayes probability was **0.0906**.

These values are very similar. The pivot table estimate is more direct because it uses the actual observed customers with CC = 1 and Online = 1. The Naive Bayes estimate uses conditional probabilities and assumes that the predictors are conditionally independent.

## G. Naive Bayes Model

```{r}
nb.model <- naiveBayes(Loan ~ CC + Online, data = train.df)

nb.model
```

```{r}
new.customer <- data.frame(
  CC = factor(1, levels = levels(train.df$CC)),
  Online = factor(1, levels = levels(train.df$Online))
)

predict(nb.model, new.customer, type = "raw")
```

The model output gives:

- Loan = 0: **0.9094**
- Loan = 1: **0.0906**

The entry for Loan = 1 corresponds to:

P(Loan = 1 | CC = 1, Online = 1)

This matches the manual Naive Bayes calculation from Part E.

## Conclusion

For a customer who owns a bank credit card and actively uses online banking services, the estimated probability of accepting the personal loan offer is about **9%**. The pivot table and Naive Bayes results are almost identical.
